Randomized trace estimation for parameter-dependent matrices applied to spectral density approximation

### 2025-02-19 paper/spectraldensity.tex
restructure Chebyshev expansion and adjust notation

To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we approximate the function 
$s\mapsto g_{\sigma}(t- s)$ by a truncated
Chebyshev expansion $g_{\sigma}^{(m)}(t-s) = 
\sum_{l=0}^{m} \mu_l(t) T_l(s)$
with the expansion coefficients $\mu_l(t) \in \mathbb R$ and the Chebyshev polynomials
$T_l(s) = \cos(l \cdot \arccos(s))$ for $l = 0,1,\ldots,m$.
Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, this gives rise to
%need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$. However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation. For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation. In every $t \in \mathbb{R}$ we expand 
\begin{equation}
    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}),
    \label{equ:matrix-expansion}
\end{equation}
\textcolor{red}{??? MOVE BELOW! This is not at a good place
For later purposes, we recall the well-known
three-term recurrence relation
\begin{equation}
    T_0(s) = 1,\quad  T_1(s) = s,\quad 
    T_l(s) = 2 s T_{l-1}(s) - T_{l-2}(s) \text{ for $l \geq 2$.}
    \label{equ:chebyshev-recurrence}
\end{equation}}
Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
\begin{equation}
    \phi_{\sigma}(t) \approx \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
    \label{equ:expanded-spectral-density}
\end{equation}
\begin{remark}
For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum of $A$~\cite{zhou-2011-bounding-spectrum}. The usual affine transform
$\tau(t) = \frac{2t - a - b}{b - a}$
then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the expansion \refequ{equ:matrix-expansion} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on, without loss of generality that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
\end{remark}

@@

\reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$. It now also allows us to bound the $L^1$-error we make when expanding the spectral density.

\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
        \label{equ:chebyshev-interpolation-spectral-density}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$. %The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
\end{theorem}

\begin{proof}
    We observe that for all $t \in [-1, 1]$
    \begin{align}
        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
        && \text{(conservative upper bound)} \notag \\
        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
    \end{align}
    Then, Hölder's inequality allows us to conclude
    \begin{equation}
        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
        \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
    \end{equation}
\end{proof}

@@

To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative. We can think of many ways in which this can be achieved: by using Jackson damping \cite{jackson-1912-approximation-trigonometric,braverman-2022-sublinear-time}, albeit at a significantly worse rate of convergence; by shifting up the expansion $g_{\sigma}^{(m)} + \rho$ with a large enough $\rho > 0$, at the loss of some of the low-rank structure of $g_{\sigma}(t \mtx{I}_n - \mtx{A})$; or by directly minimizing the approximation error over all non-negative Chebyshev polynomials \cite{fejer-1916-uber-trigonometrische}, which realistically only works for small $m$. In the end, we propose the following method to produce a non-negative Chebyshev expansion: Fix an even $m$ and expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
\begin{equation}
    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
\end{equation}
Then, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
\begin{equation}
    \underline{g}^{(m)}_{\sigma}(t - s) = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s).
    \label{equ:non-negative-chebyshev-expansion}
\end{equation}
With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive in \refsec{subsubsec:chebyshev-nystrom-analysis}. In \reftab{tab:chebyshev-timing-interpolation}, the non-negative expansion can be seen to only be slightly slower than the normal one.

\begin{table}[ht]
    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{quadrature FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and computing a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}). We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$. For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
    \label{tab:chebyshev-timing-interpolation}
   \input{tables/interpolation.tex}
\end{table}

%\todo{
%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement. Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}. The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
%}

\subsection{Computation of expansion coefficients}

\subsubsection{Fast and consistent squaring}
\label{subsubsec:dct}

??? UNTIL HERE ???
\color{black}

It is well known that the coefficients of a Chebyshev expansion of a function can --- in exact arithmetic --- be computed by inverting the discrete cosine transform (DCT)
\begin{equation}
    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
    \label{equ:discrete-cosine-transform}
\end{equation}
between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$. This can be accomplished in $\mathcal{O}(m \log(m))$ operations. \cite[Algorithm 1]{lin-2017-randomized-estimation} uses a slightly different approach, which only approximates the expansion coefficients and is noticeably slower (cf. \reftab{tab:chebyshev-timing-interpolation}).

A highly useful property of the Chebyshev expansion --- of which we will make use twice in the following pages --- is the fact that we can square such an expansion in $\mathcal{O}(m \log(m))$ operations \cite{baszenski-1997-fast-polynomial}, which the below algorithm implements.

\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
    \begin{algorithmic}[1]
        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_0, \dots, \mu_m$ and the remaining $m$ are $0$
    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_l$
    \label{lin:inverse-DCT}
    \end{algorithmic}
\end{algo}
On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.

@@

In the end, we propose the following approximation for the smoothed spectral density
\begin{equation}
    \underline{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A}))
    \label{equ:chebyshev-nystrom-formula}
\end{equation}
which is the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ \refequ{equ:nystrompp-trace-estimator} applied to the non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion}.

@@

    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\underline{\phi}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
    \State Compute $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ for all $t_i$ with non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}
    \State Compute $\{\nu_l(t_i)\}_{l=0}^{2m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}

@@

            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \underline{\mu}_l(t_i) \mtx{X}$ \Comment{assemble $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \underline{\mu}_l(t_i) \mtx{Y}$ \Comment{assemble $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
            \State $\ell(t_i) \gets \ell(t_i) + \underline{\mu}_l(t_i) z$ \Comment{assemble $\Trace(\mtx{\Psi}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}

>>>>

To estimate the trace in \refequ{equ:smooth-spectral-density} efficiently, we approximate the function 
$s\mapsto g_{\sigma}(t- s)$ by a truncated
Chebyshev expansion $g_{\sigma}^{(m)}(t-s) = 
\sum_{l=0}^{m} \mu_l(t) T_l(s)$
with the expansion coefficients $\mu_l(t) \in \mathbb R$ and the Chebyshev polynomials
$T_l(s) = \cos(l \cdot \arccos(s))$ for $l = 0,1,\ldots,m$.
Assuming that the spectrum of $\mtx{A}$ is contained in $[-1,1]$, this gives rise to
%need to evaluate the involved matrix function $g_{\sigma}(t\mtx{I}_n - \mtx{A})$. However, doing this exactly would require us to first diagonalize $\mtx{A}$, which is often a prohibitively expensive operation. For algorithmic reasons (see \refsec{subsubsec:chebyshev-nystrom-implementation}), we seek for approximations of this matrix function in terms of affine linear expansions, of which the was determined to be most suitable due to its approximation error guarantees and ease of manipulation. In every $t \in \mathbb{R}$ we expand 
\begin{equation}
    g_{\sigma}(t\mtx{I}_n - \mtx{A}) \approx g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) = \sum_{l=0}^{m} \mu_l(t) T_l(\mtx{A}),
    \label{equ:matrix-expansion}
\end{equation}
Inserting~\refequ{equ:matrix-expansion} into~\refequ{equ:smooth-spectral-density} yields the approximate smoothed spectral density
\begin{equation}
    \phi_{\sigma}(t) \approx \phi_{\sigma}^{(m)}(t) =  \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})).
    \label{equ:expanded-spectral-density}
\end{equation}
\begin{remark}
For a general symmetric matrix $\mtx{A}$, one first estimates an interval $[a,b]$ containing the spectrum of $A$~\cite{zhou-2011-bounding-spectrum}. The usual affine transform
$\tau(t) = \frac{2t - a - b}{b - a}$
then lets us define a matrix $\bar{\mtx{A}} = \tau(\mtx{A})$ for which the spectrum is contained in $[-1, 1]$ and, hence, the expansion \refequ{equ:matrix-expansion} is valid.\footnote{Note that one has to adjust the smoothing parameter $\bar{\sigma} = 2 \sigma (b - a)^{-1}$ to obtain an undistorted approximation of the original spectrum when inverting the transformation.} This allows to assume, from now on, without loss of generality that the spectrum of $\mtx{A}$ is contained in $[-1, 1]$.
\end{remark}

@@

\reflem{lem:chebyshev-error} shows that when $0 < \sigma \ll 1$ and if we choose $m = \mathcal{O}(\sigma^{-1}(\log(\varepsilon^{-1}) + \log(\sigma^{-1})))$, the error we induce with the Chebyshev expansion is at most $\mathcal{O}(\varepsilon)$. It now also allows us to bound the $L^1$-error we make when expanding the spectral density.
\begin{align}
    \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t 
    &= \int_{-1}^{1} \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|~\mathrm{d}t \notag \\
    &\leq \int_{-1}^{1} \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|~\mathrm{d}t \notag \\
    &\leq 2 E_{\sigma, m}
    \label{equ:chebyshev-interpolation-spectral-density}
\end{align}

%\begin{theorem}{$L^1$-error of Chebyshev expansion of spectral density}{chebyshev-error}
%    Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the spectal density with Chebyshev expansion $\phi_{\sigma}^{(m)}$ satisfies
%    \begin{equation}
%        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 2 E_{\sigma, m}
%        \label{equ:chebyshev-interpolation-spectral-density}
%        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
%        %\label{equ:2-chebyshev-interpolation-error}
%    \end{equation}
%    for all degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and the error $E_{\sigma, m}$ from \reflem{lem:chebyshev-error}.%, and $t \in \mathbb{R}$. %The same result holds for the expansion of a non-negative Chebyshev expansion of \emph{even} degree $m \in \mathbb{N}$ but with $\underline{E}_{\sigma, m}$.
%\end{theorem}
%
%\begin{proof}
%    We observe that for all $t \in [-1, 1]$
%    \begin{align}
%        &\left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right| \notag \\
%        &= \left| \frac{1}{n} \Trace(g_{\sigma}(t\mtx{I}_n - \mtx{A})) - \frac{1}{n} \Trace(g_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A})) \right|
%        && \text{(definitions \refequ{equ:smooth-spectral-density} and \refequ{equ:expanded-spectral-density})} \notag \\
%        &= \left| \frac{1}{n} \sum_{i=1}^n \left(g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i)\right) \right|
%        && \text{($\lambda_1, \dots, \lambda_n$ eigenvalues of $\mtx{A}$)} \notag \\
%        &\leq \max_{i = 1, \dots, n} \left| g_{\sigma}(t - \lambda_i) - g_{\sigma}^{(m)}(t - \lambda_i) \right|
%        && \text{(conservative upper bound)} \notag \\
%        &\leq E_{\sigma, m}.% ~ \text{(or $\underline{E}_{\sigma, m}$)}.
%        && \text{(using \reflem{lem:chebyshev-error})} \notag \\
%    \end{align}
%    Then, Hölder's inequality allows us to conclude
%    \begin{equation}
%        \int_{-1}^{1} | \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) | ~\mathrm{d}t
%        \leq 2 \sup_{t \in [-1, 1]} \left| \phi_{\sigma}(t) - \phi_{\sigma}^{(m)}(t) \right|
%        \leq 2 E_{\sigma, m}.% ~ \text{(or $2 \underline{E}_{\sigma, m}$)}.
%    \end{equation}
%\end{proof}

@@

To still apply the bound from \refthm{thm:nystrom-pp} in this scenario, we will have to ensure that the Chebyshev expansion is non-negative. We propose the following method to produce a non-negative Chebyshev expansion \textcolor{red}{[REFERENCE NEEDED]}: Fix an even $m$ and expand the square root of the smoothing kernel in a basis of Chebyshev polynomials up to degree $m/2$
\begin{equation}
    \sqrt{g_{\sigma}}^{(\sfrac{m}{2})}(t - s) = \sum_{l=0}^{m/2} \xi_l(t) T_l(s).
    \label{equ:square-root-chebyshev-expansion}
\end{equation}
Then, square this expansion with \refalg{alg:chebyshev-squaring} to obtain the non-negative Chebyshev expansion
\begin{equation}
    \underline{g}_{\sigma}^{(m)}(t - s) = \sum_{l=0}^{m} \underline{\mu}_l(t) T_l(s).
    \label{equ:non-negative-chebyshev-expansion}
\end{equation}

??? UNTIL HERE ???

%With a complexity of $\mathcal{O}(m \log(m))$ this procedure is computationally inexpensive and allows for very similar convergence bounds to \reflem{lem:chebyshev-error}, which we will derive \textcolor{red}{in the following?} in \refsec{subsubsec:chebyshev-nystrom-analysis}.

\color{black}

%To this extent, we will first analyze the non-negative Chebyshev expansion.

The resulting non-negative version of the approximate smoothed spectral density \refequ{equ:expanded-spectral-density}
\begin{equation}
    \phi_{\sigma}(t) \approx \underline{\phi}_{\sigma}^{(m)}(t) = \frac{1}{n} \Trace(\underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}))
    \label{equ:non-negative-expanded-density}
\end{equation}
then satisfies the following bound:

\begin{lemma}{$L^1$-error of non-negative approximate smooth spectral density}{non-negative-chebyshev-error}
Let $\phi_{\sigma}$ be the smoothed spectral density of a symmetric matrix $\mtx{A} \in \mathbb{R}^{n \times n}$ whose spectrum is contained in $[-1, 1]$. Then the non-negative approximate smoothed spectral density $\underline{\phi}_{\sigma}^{(m)}$ satisfies
    \begin{equation}
        \int_{-1}^{1} \left| \phi_{\sigma}(t) - \underline{\phi}_{\sigma}^{(m)}(t) \right|~\mathrm{d}t \leq 4 \sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}} \equiv \underline{E}_{\sigma, m}
        \label{equ:chebyshev-interpolation-sup-error-kernel-nonneg}
        %\lVert  \phi_{\sigma} - \phi_{\sigma}^{(m)} \rVert _1 &\leq \frac{2\sqrt{2}}{\sigma^2} (1 + \sigma)^{-m}.
        %\label{equ:2-chebyshev-interpolation-error}
    \end{equation}
    for all \emph{even} degrees $m \in \mathbb{N}$, smoothing parameters $\sigma > 0$, and with $E_{\sigma, m}$ as defined in \reflem{lem:chebyshev-error}.
\end{lemma}

\begin{proof}
    For any numbers $a, b \in \mathbb{R}$ it holds $| a^2 - b^2 | \leq (2 | a | + | a - b |)  | a - b |$. Therefore,
    \begin{equation}
        \left| g_{\sigma} - \underline{g}_{\sigma}^{(m)} \right| = \left| (\sqrt{g_{\sigma}})^2 - (\sqrt{g_{\sigma}}^{(\sfrac{m}{2})})^2 \right| \leq \left( 2 \left| \sqrt{g_{\sigma}} \right| + \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right| \right) \left| \sqrt{g_{\sigma}} - \sqrt{g_{\sigma}}^{(\sfrac{m}{2})} \right|.
    \end{equation}
    Since $\sqrt{g_{\sigma}} = \sqrt{2 \sigma \sqrt{2 \pi}} \cdot g_{\sqrt{2}\sigma}$, we can apply \reflem{lem:chebyshev-error} with $\sigma \to \sqrt{2}\sigma$ and $m \to m/2$ along with the bound $|\sqrt{g_{\sigma}}| \leq 1/\sqrt{\sigma \sqrt{2 \pi}}$ to get
    \begin{align}
        &\sup_{t \in [-1, 1]} \left| g_{\sigma}(t - s) - \underline{g}_{\sigma}^{(m)}(t - s) \right| \notag \\
        &\leq \left( 2 \frac{1}{\sqrt{ \sigma \sqrt{2\pi}}} + \sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right)\sqrt{2 \sigma \sqrt{2 \pi}} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}} \notag \\
        &= 2\sqrt{2} \left(1 + \sigma \sqrt{\pi} \cdot E_{\sqrt{2}\sigma, \sfrac{m}{2}}\right) E_{\sqrt{2}\sigma, \sfrac{m}{2}}.
    \end{align}
    An identical treatment to \refequ{equ:chebyshev-interpolation-spectral-density} then leads to the statement of the lemma.
\end{proof}

On one hand, in the non-negative expansion of degree $m$ \refequ{equ:non-negative-chebyshev-expansion}, the underlying Chebyshev expansion -- whose error is shown in \reflem{lem:chebyshev-error} to exponentially decay with $m$ -- is only computed to half the degree (i.e. $m/2$) of the standard Chebyshev expansion, hence, usually not as accurate. On the other hand, in the non-negative expansion we approximate $\sqrt{g_{\sigma}} \propto g_{\sqrt{2}\sigma}$ which is less \enquote{pointy} than $g_{\sigma}$ and therefore easier to approximate with Chebyshev polynomials. Overall, we will need to choose the degree $m$ of the non-negative expansion roughly a factor of $\sqrt{2}$ larger to match the approximation error of the standard expansion (cf. \reffig{fig:interpolation-issue}).

%In the following, we will show how the coefficients $\underline{\mu}_0(t), \dots, \underline{\mu}_m(t)$ of the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion} can efficiently be computed and that this expansion exhibits similar convergence properties to the standard Chebyshev expansion.
%\todo{
%Maybe mention indefinite Nyström [Nakatsukasa], which won't give us desirable guarantees, and refinement of Chebyshev expansion [Francis Bach], which only slightly improves, but is much harder to implement. Side-note: it is possible to refine the coefficients of the Chebyshev expansion to get a better approximation in the supremum-norm over the set of all non-negative polynomials \cite%{fejer-1916-uber-trigonometrische}. The non-convex nature of this problem and the relatively small improvement we get with this technique made us decide to not pursue this further.
%}

%\subsection{Computation of expansion coefficients}

\subsubsection{Computation of Chebyshev coefficients}
\label{subsubsec:dct}

It is well known that the coefficients of a Chebyshev expansion of a function can --- in exact arithmetic --- be computed by inverting the discrete cosine transform (DCT)
\begin{equation}
    g_{\sigma}(t - \cos(s_i)) = \sum_{l=0}^{m} \mu_l(t) \cos\left(\frac{\pi i l}{m} \right)
    \label{equ:discrete-cosine-transform}
\end{equation}
between the expansion coefficients $\mu_l(t),~i=0,\dots,m$ and the function evaluations at the Chebyshev nodes of the second kind $s_i = \cos(\pi i / m),~i=0,\dots,m$ \cite{vanloan-1992-computational-frameworks, baszenski-1997-fast-polynomial}. This can be accomplished in $\mathcal{O}(m \log(m))$ operations.

A desirable property of the Chebyshev expansion is the fact that such an expansion can be squared in $\mathcal{O}(m \log(m))$ operations by using the DCT and its inverse \cite{baszenski-1997-fast-polynomial}. This allows us to efficiently compute the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion}, and will also be useful in \refsec{subsec:chebyshev-nystrom}. The below algorithm implements this observation.

\begin{algo}{Fast squaring of Chebyshev expansion}{chebyshev-squaring}
    \begin{algorithmic}[1]
        \Statex \textbf{Input:} Coefficients $\{ \mu_l \}_{l=0}^{m}$ of the expansion $\sum_{l=0}^{m} \mu_l T_l$
        \Statex \textbf{Output:} Coefficients $\{ \nu_l \}_{l=0}^{2m}$ such that $\sum_{l=0}^{2m} \nu_l T_l = (\sum_{l=0}^{m} \mu_l T_l)^2$
        \State Define $\vct{\mu} \in \mathbb{R}^{2m + 1}$ whose first $m + 1$ entries are the coefficients $\mu_0, \dots, \mu_m$ and the remaining $m$ are $0$
    \State Compute $\vct{f} = \operatorname{DCT}(\vct{\mu})$
    \State Compute $\vct{\nu} = \operatorname{DCT}^{-1}(\vct{f} \odot \vct{f})$ whose entries are the coefficients $\nu_0, \dots, \nu_{2m}$
    \label{lin:inverse-DCT}
    \end{algorithmic}
\end{algo}
On \reflin{lin:inverse-DCT}, $\odot$ denotes the entrywise product.

In \cite{lin-2017-randomized-estimation}, a slightly different approach to computing the expansion coefficients in \refequ{equ:discrete-cosine-transform} is taken. They express the coefficients $\mu_l(t)$ in terms of an integral, which they subsequently approximate with a quadrature supported in at least $2m + 1$ nodes. They evaluate the resulting approximation with the fast Fourier transform (FFT). Besides not yielding the exact expansion coefficients $\mu_l(t)$, this procedure is also significantly slower due to the larger number of quadrature nodes and the implementational benefits of the DCT over the FFT. In \reftab{tab:chebyshev-timing-interpolation}, we compare a benevolent implementation of \cite[Algorithm 1]{lin-2017-randomized-estimation} with our approach, both for the standard and the non-negative Chebyshev expansion.

\begin{table}[ht]
    \caption{We compare the runtime in milliseconds for computing the coefficients of the Chebyshev expansion of the smoothing kernel $g_{\sigma}$ \refequ{equ:smoothing-kernel} in three different ways: as proposed in \cite[Algorithm 1]{lin-2017-randomized-estimation} (\emph{FFT}), using the discrete cosine transform (\emph{DCT}) \refequ{equ:discrete-cosine-transform}, and using the discrete cosine transform to compute a non-negative Chebyshev expansion with \refalg{alg:chebyshev-squaring} (\emph{non-negative DCT}). We use $\sigma=0.005$, $n_t=1000$ parameter values, and various degrees $m$. For stability, we average over 7 runs of the algorithms and repeat these runs 1000 times to form the mean and standard deviation which are given in the below table.}
    \label{tab:chebyshev-timing-interpolation}
   \input{tables/interpolation.tex}
\end{table}

@@

Combining the results from \refsec{sec:analysis} for $\mtx{B}(t) = \underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})$ with the Chebyshev expansion framework from \refsec{subsec:chebyshev-expansion}, we now approximate the smoothed spectral density with
\begin{equation}
    \phi_{\sigma}(t) \approx \widetilde{\underline{\phi}}_{\sigma}^{(m)}(t) = \frac{1}{n} \Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}(\underline{g}_{\sigma}^{(m)}(t \mtx{I}_n - \mtx{A})),
    \label{equ:chebyshev-nystrom-formula}
\end{equation}
which is the Nyström++ estimator $\Nystrpp{\mtx{\Omega}}{\mtx{\Psi}}$ \refequ{equ:nystrompp-trace-estimator} applied to the non-negative Chebyshev expansion $\underline{g}_{\sigma}^{(m)}$ \refequ{equ:non-negative-chebyshev-expansion}.

@@

    \Statex \textbf{Output:} Approximate evaluations of the spectral density $\{\widetilde{\underline{\phi}}_{\sigma}^{(m)}(t_i)\}_{i=1}^{n_t}$
        \State Compute the coefficients $\{\xi_l(t_i)\}_{l=0}^{m/2}$ of the Chebyshev expansion of $\sqrt{g_{\sigma}}$ \refequ{equ:square-root-chebyshev-expansion}
        \State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev expansion by applying \refalg{alg:chebyshev-squaring} to $\{\xi_l(t_i)\}_{l=0}^{m/2}$
        \State Compute the coefficients $\{\underline{\nu}_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev expansion by applying \refalg{alg:chebyshev-squaring} to $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$
    %\State Compute the coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ of the non-negative Chebyshev expansion \refequ{equ:non-negative-chebyshev-expansion} for all points $t_i$ \textcolor{red}{maybe more details needed (maybe compute $\xi$ for m/2, then twice Alg 3.3)}
    %\State Compute the coefficients $\{\nu_l(t_i)\}_{l=0}^{2m}$ of the square of the non-negative Chebyshev expansion with coefficients $\{\underline{\mu}_l(t_i)\}_{l=0}^{m}$ using \refalg{alg:chebyshev-squaring}%\refalg{alg:2-chebyshev-chebyshev-expansion}

@@

            \State $\mtx{K}_1(t_i) \gets \mtx{K}_1(t_i) + \underline{\mu}_l(t_i) \mtx{X}$ \Comment{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Omega}$}
            \State $\mtx{L}_1(t_i) \gets \mtx{L}_1(t_i) + \underline{\mu}_l(t_i) \mtx{Y}$ \Comment{assembles $\mtx{\Omega}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi}$}
            \State $\ell(t_i) \gets \ell(t_i) + \underline{\mu}_l(t_i) z$ \Comment{assembles $\Trace(\mtx{\Psi}^{\top} \underline{g}_{\sigma}^{(m)}(t\mtx{I}_n - \mtx{A}) \mtx{\Psi})$}
